Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-11 paper/paper.tex
draft of section 1,4. Removed dead code from comments

\renewcommand{\Gamma}{\mathgtt{G}}
\renewcommand{\Delta}{\mathgtt{D}}

@@

\newcommand{\SPF}{\mathsf{SPF}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\textbf{String}}
\newcommand{\stringg}{\mathtt{String}}
\newcommand{\charg}{\mathtt{Char}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathtt{Bool}}
\newcommand{\true}{\mathtt{true}}
\newcommand{\false}{\mathtt{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$~}
\newcommand{\lnld}{$\textrm{LNL}_D$}

@@

\begin{abstract}
  We propose \theoryname~(\theoryabbv), a domain-specific dependent type theory
  for the definition of intrinsically verified parsers and verification of weak
  and strong equivalences of formal grammars. The linear types in \theoryabbv
  have a simple interpretation as formal grammars, with values of the types
  interpreted as parse trees and general terms as a form of intrinsically
  verified parse transformers, of which parsers being a special case. We
  demonstrate the expressivity of this system by showing that the indexed
  inductive linear types in this calculus can be used to encode many different
  familiar grammar formalisms, including regular and context-free grammars but
  also traces of various types of automata. We can then construct strong
  equivalences between grammar and automata formalisms and construct verified
  parsers as simple linear functional programs. All of our examples have been
  implemented in a prototype implementation of \theoryabbv as a shallow
  embedding in Agda.

  Our shallow embedding is based on a denotational semantics of \theoryabbv
  where non-linear types denote ordinary sets and linear types denote formal
  grammars. To develop this semantics compositionally, we define a notion of
  general notion of model we call a \emph{Chomsky hyperdoctrine}, a
  generalization of the Kleene algebra approach to formal language theory. In
  addition to our ``standard model'' where linear types denote formal grammars,
  we demonstrate two non-standard models. The first interprets linear types as
  semantic actions over formal grammars, suggesting future extensions of our
  calculus. The second is used to structure a logical relations argument to
  prove a canonicity metatheorem for the type theory, proving that all closed
  terms of context-free linear types normalize to a parse tree.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}

\steven{TODO adapt section 1 and the abstract to reflect the rest of the paper}
%% Outline
%% - Why a framework for parser verification
%% - what intrinsic verification gets us
%% - the secret sauce: indexed linear inductive types
%% - prototype implementation
%% - overview of "abstract" formal grammars?

Parsing data from a serialized format is one of the most common tasks
in all of computing. Accordingly, this is a quite well-studied problem
spanning theoretical computer science and linguistics with such
milestones as Chomsky's hierarchy of grammars
\cite{chomThreeModels1956}, practical algorithms for parsing of
regular and context-free grammars, and variants
\cite{KNUTH1965607,Earley1970}, as well as implementations of
practical tools for the generation of efficient parsers
\cite{Johnsonyacc}.

The theory of formal languages and parsing is one of the oldest and most
thoroughly developed area of theoretical computer science.
The central object of study is a \emph{formal language} $L$ over an alphabet
$\Sigma$, which is classically defined as a
\emph{subset} of strings $L \subseteq \String = \Sigma^{*}$. The expressivity of formal
languages is sufficient for defining a \emph{recognizer} of a language $L$:
a function $r : \String \to \Prop$ such that $r$ maps each string $w$ to the
proposition that $w$ is in the language $w \in L$. This technique of language
recognition can concisely specify lexical analysis, and even lead to verified
lexer implementations \cite{egolfVerbatim,Ouedraogo_2023}.

While there are ample theory and tooling to handle the first stage of a language frontend, the final
frontier for parsing research is the development of
\emph{formally verified} parser implementations. Due to the
ubiquitous nature of parsing, these would useful components of many
formally verified software systems: compilers and servers most
especially. However, language recognition is insufficient for specifying a
parser, and for this task much of the structured foundations from the theory of
formal languages is rendered moot. Rather than returning a proposition-valued result, a parser produces a
structured \emph{parse tree} adherent to a \emph{formal grammar}.
The research on such grammars is often performed atop an ad-hoc choice of grammar
formalism, and results are proven with respect to that formalism. Critically,
the field is lacking a unifying theory across these different formalisms, even
though many of the same
compositional reason principles appear across distinct grammar formalisms.

In this work, we aim to elucidate these similarities and
provide a mathematical theory of formal grammars for constructing verified
parsers. Our approach can be summarized with the
Curry-Howard-style slogan: ``grammars as types''; and the addendum: ``parsers as
terms''. That is, we develop a type theory with a novel
denotational semantics in a category with abstract formal grammars as
objects and \emph{parse transformers} as morphisms. We say these are
\emph{abstract} formal grammars because they are not tied to any
particular syntactic formalism such as Chomskyan grammars, regular
expressions, Lambek calculus etc. The category we choose is a rich
in structure: it is a topos as well as being monoidal closed, and these
universal properties serve as the inspiration for our type theory.

The type theory itself is a form of dependent linear-non-linear type
theory: the linear types can be interpreted as grammars, whereas the
non-linear types can be interpreted as ordinary sets. As in prior work
on dependent linear-non-linear types, the linear types are allowed to
depend on the non-linear types but not vice-versa \cite{krishnaswami_integrating_2015}.

@@

\paragraph{Contributions} This paper begins in \cref{sec:synindnotion} by
giving a universal semantic characterization of formal grammar.
From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type theory
that axiomatizes the underlying structures needed to conduct formal grammar
theory. In \cref{sec:automata} we demonstrate the usage of
\theoryname for defining parser terms and internalizing results from formal language theory.
Then in \cref{sec:categorify}, we characterize the abstract categorical
models of our syntax. Afterwards in \cref{sec:othermodels}, we demonstrate several models,
including one to prove a metatheoretic canonicity theorem.
Our contributions are then:
%
\begin{itemize}
  \item A syntax-independent notion of grammar given by the category $\Grammar$
  \item \theoryname: A dependent
    linear-non-linear type theory meant to syntactically capture the latent structures
    in $\Grammar$ that enable parsing
  \item Categorical structures that capture the concepts integral to formal grammar theory, and that are notably modeled by $\Grammar$
  \item An implementation of the language, which is embedded shallowly in Agda
\end{itemize}

\subsection{A Syntax-Independent Notion of Syntax}
\label{sec:synindnotion}

The notion of formal language is central to the theory of parsing.
This definition is
especially useful as it gives a semantics to formal grammars that is completely
independent of any particular syntactic grammar formalism. Any new
notion of grammar can be given a language semantics and it provides a precise
mathematical specification for implementing a \emph{recognizer} of a language.
In this section we propose a \emph{proof-relevant} generalization of formal
languages that can accommodate richer aspects of formal language theory such as
parsing. We begin by proposing a presheaf approach to grammars that encompasses
other classes of grammars in a syntax-independent way.

\paragraph{From Formal Languages to Formal Grammars}

Our first contribution is a novel and syntax-independent characterization of
formal grammar:

\begin{definition}
  \label{def:grammar}
  A \emph{formal grammar} $G$ over a fixed alphabet $\Sigma$ is a function
  $G : \String \to \Set$.
\end{definition}

\steven{TODO need to go through and ensure that $\Sigma^{*}$ and $\String$ are each used appropriately}

We say that a grammar $G$ associates to every string $w$ the set\footnote{We
  make little comment as to which foundations are used to encode $\Set$. Our
  construction is polymorphic in choice of a proper class of small sets, a
  Grothendieck universe, a type theoretic universe, or any similar foundation.} $G w$ of parse
trees showing that $w$ matches $G$. This definition of formal grammar serves as
a mathematical specification for a parser --- taking on the same role as
languages do for recognizers. In this sense, our new characterization of
grammars is just a proof-relevant generalization of the existing notion of
formal language.

By ranging over all input strings, notice that each grammar induces a language.
For each formal grammar $G$, define the language of accepted strings $L_{G}$ as,

\[ L_{G} = \{ w \in \String ~|~ G w~\text{is inhabited} \} \]

As they are simply sets, formal languages are naturally endowed with a partial
order via subset inclusion. Just, formal grammars naturally coalesce into a
\emph{category}.

Define $\Grammar$ to be the category of formal grammars. Whose
objects are grammars, and whose morphisms are functions $f : G \to H$ between two
grammars presented by a family of functions $f^{w} : G w \to H w$ for
$w : \String$. Intuitively, we interpret these morphisms as parse
\emph{transformers} --- as constructive translations of $G$-parses into $H$-parses. Throughout this paper, we will make frequent reference to this categorical
structure on $\Grammar$.

\steven{Above sounds awkward. I want to additionally say something here that signposts that we will give more
  information about that structure without just listing it all here. }

\paragraph{From Grammars to Formal Grammars}
The category-theoretic framework can be further used to describe and compare
different notions of formal grammar. For instance, Chomskyan generative
grammars, semi-Thue systems, Montague grammars, and so on are all distinct syntactic
presentations of the same underlying idea of an abstract specification for
parsing. Each of these is an instance of a general \emph{notion of formal
  grammar} --- a category paired with a functor into
$\Grammar$. That is, they may all be interpreted in our category of grammars. In
this sense, $\Grammar$ provides an
all-encompassing semantic domain for parsing.

Upon further inspection, $\Grammar$ may equivalently
be described as the functor category $\Set^{\String}$ --- where the
set $\String$ is viewed as a discrete\footnote{Objects in $\String$
comprise the set of all strings, and the only morphisms are the
identity morphisms.} category. In fact, in this paper we will take
$\Grammar$ to be defined as the functor category $\Set^{\String}$. Such functor
categories in $\Set$, called presheaf categories, carry a
remarkable amount of structure.  Indeed they are a monoidal topos and,
as such, are a model to a substructural dependently typed theory.

Thus we
present \theoryname (\theoryabbv) as a domain-specific type theory that
abstracts over the structure of $\Grammar$. To begin, let us look at what some
programs in \theoryabbv look like.

\steven{TODO Reword this transition into the examples}

% \max{this section kind of trails off at the end with a very vague aside about a $2$-category of grammar formalisms. Maybe this should go in future work/discussion? What is the message of this section? How does it tie in with the message of the paper? What even is the message of this paper?}

% \pedro{Would building on this proposed structure make the message of the section clearer? In this case
%   we would have to be a bit clearer in terms of how familiar notions of grammar can be seen as categories with functor into $\Grammar$.}

@@

\max{was this paragraph supposed to be deleted?}  Next we present the
full details of the syntax and typing of \theoryabbv. The syntax of
\theoryabbv is based on the dependent linear-non-linear (\lnld)
calculus from \cite{dlnl}. As in \lnld, \theoryabbv includes both
non-linear dependent types, as well as linear types, which are allowed
to depend on the non-linear types, but not on other linear types.
%
The main point of departure from \lnld's design in that the linear
typing is \emph{non-commutative}, i.e., that exchange is not an
admissible structural rule. Furthermore, we add a general-purpose
indexed inductive linear type connective, as well as an
\emph{equalizer} type, which we will show allows us to perform
inductive proofs of equalities between linear terms.
%
Finally, we add base types and axioms to the system specifically to
model formal grammars and parsing.

@@

Next, we introduce one of the most complex and important linear
type constructors of our development, \emph{indexed inductive linear
types} $\mu F$. To encode these, we first add a non-linear type
$\SPF\,X$ of \emph{strictly-positive functorial} linear type
expressions indexed by a non-linear type $X$. We give the rules for
strictly-positive functors in \Cref{fig:spf}. First, we add operations
for a recursive reference $\Var{x}$ the constant functor, $K$, tensor
products and additive conjunction and disjunction. As an example,
\max{TODO: make an example}.  Next we add an operation $\el(F)$ which
interprets our syntactic codes as an actual type constructor, and
define it for each of the operations given above. Lastly, we give a
\emph{functorial action} $\map(F)$ for each of the codes.\max{TODO:
  find a citation for this kind of approach. Container papers work
  this way for instance. I think there is a linear container paper
  that uses this approach}

With this notion of strictly-positive functor so defined, in
\Cref{fig:iilt} we next add a linear type constructor $\mu F$, which
defines a family of mutually-recursive inductive types for the family
of type expressions $F$. The introduction rule for this is
$\mathsf{roll}$, which constructs an element of $\mu F\,x$ from its
one-level unfolding. The elimination principle is defined by a
$\mathsf{fold}$ operation. The equations ensure that this makes the
family an \emph{initial algebra} for the type expressions. That is the
$\beta$ rule says that a $\mathsf{fold}$ applied to a $\mathsf{roll}$
is equivalent to $\mathsf{map}$ping the $\mathsf{fold}$ over all the
sub-expressions, which means that $\mathsf{fold}$ is an algebra
homomorphism from $\mathsf{roll}$ to $\mathsf{f}$. Then the $\eta$
rule says that $\mathsf{fold}$ is the \emph{unique} such homomorphism.

@@

In \cref{fig:nfatrace} we define a linear type of traces
through an NFA $\N$. $\Trace_{\N}$ is an inductive type indexed by the first
state $\s : \N.\states$ of a trace and a Boolean $b : \Bool$ that marks if the
trace ends in an accepting state. Traces in $\N$ may be built through one of three constructors. We
may terminate a trace at an accepting state with the constructor $\nil$. If we
had a trace beginning at the destination state of a transition, then we may use
the $\cons$ constructor to linearly combine that trace with a parse of the label
of the transition to build a trace beginning at the source of the transition.
Finally, if we had a trace beginning at the destination of an
$\epsilon$-transition then we may use $\epscons$ to pull it back along the
$\epsilon$-transition and construct a trace beginning at the source of the $\epsilon$-transition.

\begin{figure}
\begin{lstlisting}
data Trace(*@$_{\N}$@*) : (s : N.states) (b : Bool) (*@$\to$@*) L where
  nil : (*@$\uparrow$@*)(&[ s (*@$\in$@*) N.states ] Trace(*@$_{\N}$@*)(s , N.isAcc s))
  cons : (*@$\uparrow$@*)(&[ t (*@$\in$@*) N.transitions ] &[ b (*@$\in$@*) Bool ]
           ('N.label(t)' (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.dst(t) , b) (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.src(t) , b)))
  (*@$\epsilon$@*)cons : (*@$\uparrow$@*)(&[ t (*@$\in$@*) N.(*@$\epsilon$@*)transitions ] &[ b (*@$\in$@*) Bool ]
           (Trace(*@$_{\N}$@*)(N.(*@$\epsilon$@*)dst(t) , b) (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.(*@$\epsilon$@*)src(t) , b)))
\end{lstlisting}
\caption{Traces through an NFA $N$ as an indexed inductive type}
\label{fig:nfatrace}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}
data Trace(*@$_{\D}$@*) : (s : D.states) (b : Bool) (*@$\to$@*) L where
  nil : (*@$\uparrow$@*)(&[ s (*@$\in$@*) D.states ] Trace(*@$_{\D}$@*)(s , D.isAcc s))
  cons : (*@$\uparrow$@*)(&[ c (*@$\in$@*) (*@$\Sigma$@*) ] &[ s (*@$\in$@*) D.states ] &[ b (*@$\in$@*) Bool ]
             ('c' (*@$\lto$@*) Trace(*@$_{\D}$@*)(D.(*@$\delta$@*) c s , b) (*@$\lto$@*) Trace(*@$_{\D}$@*)(s , b)))
\end{lstlisting}
\caption{Traces through the DFA $D$ as an indexed inductive type}
\label{fig:dfatrace}
\end{figure}

@@

Because DFAs are deterministic, we are able to prove that their type of traces
are unambiguous and define a parser for them. In particular we show that
$\LinPiTy {s} {D.\states} {\LinSigTy {b} {\Bool} \Trace_{D}(s , b)}$ is strongly
equivalent to
$\stringg$, which simultaneously constructs a parser and demonstrates that
traces through $D$ is unambiguous. Intuitively this strong equivalence says that we may
unambiguously build traces through $D$ beginning at any state $s$, and further
that from any trace through $D$ we can uniquely recover the string that built
the trace.

The first half of the equivalence,
$\parse_{D} : \ltonl {\left( \stringg \lto \LinPiTy {s} {D.\states} {\Trace_{D}(s)} \right)}$,
is defined by induction on strings in \cref{fig:dfaparser}. If this string is empty, then $\parse_{D}$
defines a linear function that terminates a trace at the input state $s$. If the
string is nonempty, then $\parse_{D}$ walks forward from the input state $s$ by
the character at the head of the string.

\begin{figure}
\begin{lstlisting}
parse : (*@$\uparrow$@*)(String (*@$\lto$@*) &[ s (*@$\in$@*) D.states ] (*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*)(s , b))
parse String.nil s = (*@$\sigma$@*) (D.isAcc s) (Trace(*@$_{\D}$@*).nil s)
parse (String.cons ((*@$\sigma$@*) c a) w) s = let (*@$\sigma$@*) b t = parse w (D.(*@$\delta$@*) c s) in
                           (*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c s b a t)
\end{lstlisting}
\caption{Parser for DFA traces}
\label{fig:dfaparser}
\end{figure}

The second half of the equivalence,
$\print_{D} : \ltonl {\left( \LinPiTy {s} {D.\states} {\Trace_{D}(s)} \lto \stringg \right)}$,
is given in \cref{fig:printdfa} by induction on traces. If the trace is defined
via $\nil$, then $\print_{D}$ returns the empty string. Otherwise, if the trace
is defined by $\cons$ then $\parse_{D}$ appends the character $c$ to the output
string and recurses.

\begin{figure}
\begin{lstlisting}
print : (s : D.states) (*@$\to$@*) (*@$\uparrow$@*)(((*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*)(s , b)) (*@$\lto$@*) String)
print s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).nil .s)) = String.nil
print s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c (D.(*@$\delta$@*) c .s) b a trace)) =
  String.cons ((*@$\sigma$@*) c a) (print (D.(*@$\delta$@*) c s) ((*@$\sigma$@*) b trace))
\end{lstlisting}
% \[
% \inferrule
% {
% }
% {tr : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool } {\Trace_{D}(s, b)}} \vdash asdf : \stringg}
% \]
\caption{Printer for DFA traces}
\label{fig:printdfa}
\end{figure}

>>>>

\renewcommand{\Gamma}{\mathgtt{G}}
\renewcommand{\Delta}{\mathgtt{D}}
\renewcommand{\Sigma}{\mathgtt{S}}
\renewcommand{\mu}{\mathgtt{m}}

@@

\newcommand{\SPF}{\texttt{SPF}}
\newcommand{\Var}{\texttt{Var}}
\newcommand{\K}{\texttt{K}}
\newcommand{\map}{\texttt{map}}
\newcommand{\roll}{\texttt{roll}}
\newcommand{\fold}{\texttt{fold}}
\newcommand{\inl}{\texttt{inl}}
\newcommand{\inr}{\texttt{inr}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\textbf{String}}
\newcommand{\stringg}{\texttt{String}}
\newcommand{\charg}{\mathtt{Char}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathtt{Bool}}
\newcommand{\true}{\mathtt{true}}
\newcommand{\false}{\mathtt{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$~}
\newcommand{\lnld}{$\textrm{LNL}_D$}

@@

\begin{abstract}
  We propose \theoryname~(\theoryabbv), a domain-specific dependent type theory
  for the definition of intrinsically verified parsers and verification of weak
  and strong equivalences of formal grammars. The linear types in \theoryabbv
  have a simple interpretation as formal grammars, with values of the types
  interpreted as parse trees and general terms as a form of intrinsically
  verified parse transformers, of which parsers being a special case. We
  demonstrate the expressivity of this system by showing that the indexed
  inductive linear types in this calculus can be used to encode many different
  familiar grammar formalisms, including regular and context-free grammars but
  also traces of various types of automata. We can then construct strong
  equivalences between grammar and automata formalisms and construct verified
  parsers as simple linear functional programs. All of our examples have been
  implemented in a prototype implementation of \theoryabbv as a shallow
  embedding in Agda.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
\steven{This intro sounds reallchoppy but follows the following outline.

  \begin{itemize}
  \item parsing correctness is imporant because parsing is
  everywhere and failure
          could even invalidate other proof efforts.
  \item verified parsers should be
          easier to write
  \item Conal Elliott gives the right idea for what the foundation
          of parser verification should be --- type theory

  \end{itemize}
          }
Formal verification encompasses a broad suite of techniques one may use to
mathematically guarantee the correctness of software systems. We can only
reap the benefits of
any proof effort if its underlying assumptions hold. For instance, verifying the
functional correctness of an algorithm is no use if it is run on buggy hardware.
Likewise, the promised guarantees of a program may be jeopardized if it is
parsed incorrectly.

Large verified developments, like CompCert \cite{leroy_formal_2009},
invest enormous amounts of person-hours into proof development. In early
versions of CompCert, the apparent
assurances associated with their effort could indeed be invalidated through
parser bugs \cite{yangFindingUnderstandingBugs}. Parsing is a ubiquitous task
throughout computing and yet parsers are frequently unverified, even in
otherwise verified systems. The CompCert team later
developed a method for validating the correctness of LR parsers
\cite{jourdanValidatingLRParsers2012} and then replaced their parser with one
that is certified to be correct. Despite their success in  validating their C parser and other work in the
generation of  verified parsers
\cite{lasserCoStarVerifiedALL2021,firsovCertifiedNormalizationContextFree2015,danielssonTotalParserCombinators2010},
it is still arduous and highly nontrivial to write and verify a
parser. Further, distinct approaches to generating verified parsing
implementations are often relative to a different grammar formalism. We believe
that these sepearate grammar formalisms can all be unified into a
syntax-independent notion of grammar, and that this notion of grammar can be
axiomatized into a syntax that is useful for building verified parsers.

\steven{I don't like how I make the cited work sound bad above}

Elliott \cite{elliottSymbolicAutomaticDifferentiation2021}
notes that grammars can be described as type-level predicates on strings that prove
language membership. That is, a \emph{formal grammar} $A$ is a function
$\String \to \Set$ such that for a string $w$, $A~w$ is the set of proofs
showing that $w$ belongs to the language recognized by $A$. We adopt this
definition of a formal grammar.

In this paper, we present \theoryname (\theoryabbv) as a domain specific
type theory for building verified parsers. The type theory itself is a form of dependent linear-non-linear type
theory: the linear types can be interpreted as grammars, whereas the
non-linear types can be interpreted as ordinary sets. As in prior work
on dependent linear-non-linear types, the linear types are allowed to
depend on the non-linear types but not vice-versa \cite{krishnaswami_integrating_2015}.

@@

\paragraph{Contributions} This paper begins in \cref{sec:type-theory-examples}
by giving small examples of programs in \theoryabbv so that we may build intuition.
From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type theory
that axiomatizes the underlying structures needed to conduct formal grammar
theory. In \cref{sec:applicatoins} we demonstrate the
applicablity of \theoryabbv for defining parser terms and representing results
from formal language theory as proof terms in our calculus.
Then in \cref{sec:semantics-and-metatheory}, we give a denotational semantics
that makes precise the connection between \theoryabbv syntax and formal grammars.

Our contributions are then:
%
\begin{itemize}
  \item \theoryname (\theoryabbv): A dependent
        linear-non-linear type theory for building verified parsers
  \item An implementation of \theoryabbv, which is embedded shallowly in Cubical
        Agda. Throughout this paper, we will use \agdalogo~ to mark results that
        have been mechanized in our Agda development.
\end{itemize}

@@

The syntax of
\theoryabbv is based on the dependent linear-non-linear (\lnld)
calculus from \cite{krishnaswami_integrating_2015}. As in \lnld, \theoryabbv includes both
non-linear dependent types, as well as linear types, which are allowed
to depend on the non-linear types, but not on other linear types.
%
The main point of departure from \lnld's design in that the linear
typing is \emph{non-commutative}, i.e., that exchange is not an
admissible structural rule. Furthermore, we add a general-purpose
indexed inductive linear type connective, as well as an
\emph{equalizer} type, which we will show allows us to perform
inductive proofs of equalities between linear terms.
%
Finally, we add base types and axioms to the system specifically to
model formal grammars and parsing.

@@

Next, we introduce one of the most complex and important linear
type constructors of our development, \emph{indexed inductive linear
types} $\mu F$. To encode these, we first add a non-linear type
$\SPF\,X$ of \emph{strictly-positive functorial} linear type
expressions indexed by a non-linear type $X$. We give the rules for
strictly-positive functors in \Cref{fig:spf}. First, we add operations
for a recursive reference $\Var~x$ the constant functor, $K$, tensor
products and additive conjunction and disjunction. As an example,
\max{TODO: make an example}.  Next we add an operation $\el(F)$ which
interprets our syntactic codes as an actual type constructor, and
define it for each of the operations given above. Lastly, we give a
\emph{functorial action} $\map(F)$ for each of the codes.\max{TODO:
  find a citation for this kind of approach. Container papers work
  this way for instance. I think there is a linear container paper
  that uses this approach}

With this notion of strictly-positive functor so defined, in
\Cref{fig:iilt} we next add a linear type constructor $\mu F$, which
defines a family of mutually-recursive inductive types for the family
of type expressions $F$. The introduction rule for this is
$\roll$, which constructs an element of $\mu F\,x$ from its
one-level unfolding. The elimination principle is defined by a
$\fold$ operation. The equations ensure that this makes the
family an \emph{initial algebra} for the type expressions. That is the
$\beta$ rule says that a $\mathsf{fold}$ applied to a $\mathsf{roll}$
is equivalent to $\mathsf{map}$ping the $\mathsf{fold}$ over all the
sub-expressions, which means that $\mathsf{fold}$ is an algebra
homomorphism from $\mathsf{roll}$ to $\mathsf{f}$. Then the $\eta$
rule says that $\mathsf{fold}$ is the \emph{unique} such homomorphism.

@@

In \cref{fig:nfatrace} we define a linear type of traces
through an NFA $\N$. $\Trace_{\N}$ is an inductive type indexed by the first
state $\s : \N.\states$ of a trace and a Boolean $b : \Bool$ that marks if the
trace ends in an accepting state. Traces in $\N$ may be built through one of three constructors. We
may terminate a trace at an accepting state with the constructor $\nil$. If we
had a trace beginning at the destination state of a transition, then we may use
the $\cons$ constructor to linearly combine that trace with a parse of the label
of the transition to build a trace beginning at the source of the transition.
Finally, if we had a trace beginning at the destination of an
$\epsilon$-transition then we may use $\epscons$ to pull it back along the
$\epsilon$-transition and construct a trace beginning at the source of the
$\epsilon$-transition. As a shorthand, write $Parse_{N}$ for the accepting
traces out of $N.init$.

\begin{figure}
\begin{lstlisting}
data Trace(*@$_{\N}$@*) : (s : N.states) (b : Bool) (*@$\to$@*) L where
  nil : (*@$\uparrow$@*)(&[ s (*@$\in$@*) N.states ] Trace(*@$_{\N}$@*) s (N.isAcc s))
  cons : (*@$\uparrow$@*)(&[ t (*@$\in$@*) N.transitions ] &[ b (*@$\in$@*) Bool ]
           ('N.label t' (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.dst t) b (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.src t) b))
  (*@$\epsilon$@*)cons : (*@$\uparrow$@*)(&[ t (*@$\in$@*) N.(*@$\epsilon$@*)transitions ] &[ b (*@$\in$@*) Bool ]
           (Trace(*@$_{\N}$@*) (N.(*@$\epsilon$@*)dst t) b (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.(*@$\epsilon$@*)src t) b))
\end{lstlisting}
\caption{Traces through an NFA $N$ as an indexed inductive type}
\label{fig:nfatrace}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}
data Trace(*@$_{\D}$@*) : (s : D.states) (b : Bool) (*@$\to$@*) L where
  nil : (*@$\uparrow$@*)(&[ s (*@$\in$@*) D.states ] Trace(*@$_{\D}$@*) s D.isAcc s)
  cons : (*@$\uparrow$@*)(&[ c (*@$\in$@*) (*@$\Sigma$@*) ] &[ s (*@$\in$@*) D.states ] &[ b (*@$\in$@*) Bool ]
             ('c' (*@$\lto$@*) Trace(*@$_{\D}$@*) (D.(*@$\delta$@*) c s) b (*@$\lto$@*) Trace(*@$_{\D}$@*) s b))
\end{lstlisting}
\caption{Traces through the DFA $D$ as an indexed inductive type}
\label{fig:dfatrace}
\end{figure}

@@

Because DFAs are deterministic, we are able to prove that their type of traces
are unambiguous and define a parser for them. In particular we show that
$\LinPiTy {s} {D.\states} {\LinSigTy {b} {\Bool} \Trace_{D}~s~b}$ is strongly
equivalent to
$\stringg$, which simultaneously constructs a parser and demonstrates that
traces through $D$ is unambiguous. Intuitively this strong equivalence says that we may
unambiguously build traces through $D$ beginning at any state $s$, and further
that from any trace through $D$ we can uniquely recover the string that built
the trace.

The first half of the equivalence,
$\parse_{D}$,
is defined by induction on strings in \cref{fig:dfaparser}. If this string is empty, then $\parse_{D}$
defines a linear function that terminates a trace at the input state $s$. If the
string is nonempty, then $\parse_{D}$ walks forward in $D$ from the input state $s$ by
the character at the head of the string.

\begin{figure}
\begin{lstlisting}
parse(*@$_D$@*) : (*@$\uparrow$@*)(String (*@$\lto$@*) &[ s (*@$\in$@*) D.states ] (*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*) s b)
parse(*@$_D$@*) String.nil s = (*@$\sigma$@*) (D.isAcc s) (Trace(*@$_{\D}$@*).nil s)
parse(*@$_D$@*) (String.cons ((*@$\sigma$@*) c a) w) s =
   let (*@$\sigma$@*) b t = parse w (D.(*@$\delta$@*) c s) in
   (*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c s b a t)
\end{lstlisting}
\caption{Parser for DFA traces}
\label{fig:dfaparser}
\end{figure}

The second half of the equivalence,
$\print_{D}$,
is given in \cref{fig:printdfa} by induction on traces. If the trace is defined
via $\nil$, then $\print_{D}$ returns the empty string. Otherwise, if the trace
is defined by $\cons$ then $\parse_{D}$ appends the character from the most
recent transition to the output
string and recurses.

\begin{figure}
\begin{lstlisting}
print(*@$_D$@*) : (s : D.states) (*@$\to$@*) (*@$\uparrow$@*)(((*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*) s b) (*@$\lto$@*) String)
print(*@$_D$@*) s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).nil .s)) = String.nil
print(*@$_D$@*) s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c (D.(*@$\delta$@*) c .s) b a trace)) =
  String.cons ((*@$\sigma$@*) c a) (print(*@$_D$@*) (D.(*@$\delta$@*) c s) ((*@$\sigma$@*) b trace))
\end{lstlisting}
\caption{Printer for DFA traces}
\label{fig:printdfa}
\end{figure}

\steven{TODO make the Agda logo clickable, and mention its meaning somewhere}
\newcommand{\Agda}{\agdalogo}
\begin{theorem}[Determinization, \Agda]
  \label{thm:determinization}
  Given an NFA $N$, there exists a DFA $D$ such that
  $Parse_{N}$ is weakly equivalent to $Parse_{D}$.
\end{theorem}
\newcommand{\X}{\texttt{X}}
\newcommand{\Y}{\texttt{Y}}
\newcommand{\x}{\texttt{x}}
\newcommand{\y}{\texttt{y}}
\begin{proof}
  We present a variant of Rabin and Scott's classic powerset construction \cite{rabinFiniteAutomataTheir1959}. First, we build
  the DFA $\D$. Define the
  states of $\D$ to be the $\mathbb{P}_{\varepsilon}(\N.\states)$ --- the type
  of $\epsilon$-closed\footnote{A subset of states $\X$ is $\epsilon$-closed if
    for every $\s \in \X$ and $\epsilon$-transition $\s \overset{\epsilon}{\to} \s'$ we have $\s' \in \X$.} subsets of
  $\N.\states$. A subset is accepting in $\D$ if it contains an accepting state
  from $\N$. Construct the initial state of $\D$ as the
  $\epsilon$-closure of $\N.\init$. Lastly define the transition function
  of $\D$ to send the subset $X$ under the character $\c$ to the
  $\epsilon$-closure of all the states reachable from $X$ via a transition
  labelled with the character $\c$.
